---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# package.nicole.hw4

<!-- badges: start -->
[![R-CMD-check.yaml](https://github.com/nicole-jerome/package.nicole.hw4.repo/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/nicole-jerome/package.nicole.hw4.repo/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/nicole-jerome/package.nicole.hw4.repo/graph/badge.svg)](https://app.codecov.io/gh/nicole-jerome/package.nicole.hw4.repo)
<!-- badges: end -->

## Overview
The goal of package.nicole.hw4 is to recreate a mini version of tidyverse's dplyr. However, this package is written using base R commands. package.nicole.hw4 contains two functions. "filter_rows()" is similar to dplyr::filter(), and "select_columns()" is similar to dplyr::select(). This package was developed as a homework assignment for BIOSTAT625 at the University of Michigan. Go Blue!

## Installation

You can install the development version of package.nicole.hw4 from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("nicole-jerome/package.nicole.hw4.repo")
```

## Example Usage

This example outlines the functionality of filter_rows():

```{r example}
#delete when publishing
devtools::load_all()

#instead load this (uncomment following line with publishing)
#library(package.nicole.hw4)


```

This example outlines the functionality of select_columns():
```{r example2}
#delete when publishing
devtools::load_all()

#instead load this (uncomment following line with publishing)
#library(package.nicole.hw4)


```


## Getting Help

Help documentation is included for all functions, and the package vignette displays example use cases. The main place to suggest improvements and bug fixes with package.nicole.hw4 is via GitHub pull requests. Alternatively, you may reach out to the developer directly, Nicole Jerome, at ndea@umich.edu - please allow two business days for response time. 
