---
title: "mini_dplyr_vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mini_dplyr_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction
 
The package.nicole.hw4 package provides simple tools for data
manipulation using only base R. It aims to replicate tidyverse::dplyr.

It includes two main functions:

`filter_rows()` — subsets rows of data frames by logical conditions

`select_columns()` — subsets data frames by including or excluding specific columns

A third function, `df_check()`, is used internally to ensure that inputs
are valid data frames.

This vignette demonstrates how to apply these functions using the built-in
mtcars dataset.

<br>

# The mtcars dataset

mtcars is one of R's built-in datasets. It contains a row names index that indicate's the car's make/model, 11 columns of data, and 32 observations (rows). 

 Lets view the first few rows:
```{r setup}
head(mtcars)
```
We will use this dataset to illustrate the use of the `filter_rows()` and `select_columns()` functions. 

<br>

# Filtering rows with `filter_rows()`

The `filter_rows()` function allows users to subset a data frame based on a logical condition.
The condition is evaluated within the data frame, so column names can be used directly.

## Example 1: Keep cars with mpg > 20
```{r}
library(package.nicole.hw4)

filter_rows(mtcars, mpg > 20)
```

This returns only the cars whose miles per gallon (mpg) exceed 20.
<br>

## Example 2: Keep cars with < 6 cylinders
```{r}
filter_rows(mtcars, cyl < 6)
```

This returns only the cars that have less than 6 cylinders.
<br>

## Example 3: Combining mulitple conditions
You can combine logical expressions using &, |, and !.

Example:
Cars with mpg > 20 and < 6 cylinders:
```{r}
filter_rows(mtcars, mpg > 20 & cyl < 6)
```

This returns only the cars with over 20 mpg AND less than 6 cylinders.

<br>

# Selecting columns with `select_columns()`

The `select_columns()` function allows users to subset a data frame by providing columns to either include or exclude. Selecting columns by name and index are both supported. 

The `columns` argument indicates columns to include, whereas the `exclude` argument indicates columns to exclude. `select_columns()` only handles input into either `columns` or `exclude`, not both.

## Example 1: Select columns to keep by name
```{r}
select_columns(mtcars, columns = c("mpg", "hp", "wt"))
```

This returns only the columns mpg, hp, and wt, and discards the other 8.
<br>

## Example 2: Select columns to keep by index
```{r}
select_columns(mtcars, columns = c(1, 4, 6))
```

This returns the same columns as example 1, but calls them by index rather than name. 
<br>

## Example 3: Exclude columns by name
```{r}
select_columns(mtcars, exclude = c("cyl", "disp", "drat", "qsec", "vs", "am", "gear", "carb"))
```

This returns the same columns as example 1, but uses column names to indicate which columns to exclude.
<br>

## Example 4: Exclude columns by index
```{r}
select_columns(mtcars, exclude = c(2, 3, 5, 7, 8, 9, 10, 11))
```

This returns the same columns as example 1, but uses indices to indicate which columns to exclude. 

<br>

# Combining filtering and selecting

Because both functions return a regular data frame, they can be chained (sequentially) using standard R syntax.
For example:
```{r}
result <- filter_rows(mtcars, mpg > 25)
select_columns(result, columns = c("mpg", "cyl", "wt"))
```

Or, functions can be combined in a single line.
For example:
```{r}
select_columns(
  filter_rows(mtcars, mpg > 25),
  columns = c("mpg", "cyl", "wt")
  )
```

<br>

# Internal helper: `df_check()`
Although not exported, the package uses a helper function called `df_check()` to validate that the user input is a data frame. This ensures that meaningful error messages are returned if the input is invalid.

<br>

# Summary

The package.nicole.hw4 package provides:

 - A simple and intuitive row filtering function: `filter_rows()`
 
 - A flexible column inclusion/exclusion function: `select_columns()`
 
 - Validation of input data and clean error handling through `df_check`
 
 
These functions mimic the basic features of `dplyr` but are built in base R -- keeping them simple, lightweight, and easy to understand. package.nicole.hw4 is a useful tool for manipulating data intuitively. Happy data manipulating!
 
